---
title: "Random Forest in R"
author: "Victoria Scholl"
date: "4/8/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = FALSE)

library(dplyr)
library(ggplot2)
library(randomForest)
library(knitr)
library(devtools)
library(geoNEON)
library(sp)           
library(swfscMisc)    
library(rgdal)
library(rgeos)
library(taxize)
library(stringr)
library(caret)
library(reshape)
library(gridExtra)

# set working directory to neon-veg in local environment
#setwd("~/github/neon-veg")

# path to directory containing NEON l1 Woody Vegetation data 
main_path <- "~/Documents/NEON/SJER/NEON_struct-woody-plant/" 

# specify output directory path and filename of output shapefile to be written
out_dir <- "~/github/random-forest/output/"

# load local functions written in external R files. 
# source("locate_woody_veg.R")
# source("woody_df_to_shp.R")
# source("merge_vst_tables.R")
# source("get_vst_crs.R")
# source("list_tiles_with_plants.R")
# source("apply_area_threshold.R")
# source("polygon_overlap.R")
# source("get_poly.R")
source("~/github/neon-veg/check_create_dir.R")
```


## NEON woody vegetation structure data product 

The following information is available in the [NEON User Guide to Woody Plant Vegetation Structure (NEON.DP1.10098)](http://data.neonscience.org/documents/10179/2237401/NEON_vegStructure_userGuide_vA/b390b73d-b259-46d8-9522-58fd72f12f8e):

Trees >10 cm diameter are mapped are mapped as points (offsets from a reference pointID; see table vst_mapping); small diameter trees and shrubs are not mapped but are iden - fied according to the subplot or nested subplot in which they are located (see tables vst_apparen ndividual and vst_shrubgroup) as are non-woody plants (see table nst_perinidividual). 

The individualID (NEON.PLA.D##.######) is a linking variable that  es vegeta on structure measurements and
associated metadata to the following associated data products:

The Woody Plant Vegetation Structure data product is comprised of four related tables:

**vst_perplotperyear** - One record is created for each plotID visited at a site during a given eventID. Recorded metadata include presence/absence of various woody growth forms, and nestedSubplotArea used to measure each growth form (if applicable).

**vst_mappingandtagging** - One or more records expected for each tagged woody individual (individualID) for all  me. A subset of tagged individuals are also mapped. Corrections to taxonID or mapped loca on through  me will result in duplicate individualIDs. In the case of duplicates, users are advised to retain the record with the latest date. Non-woody individuals that are also mapped and/or tagged (e.g., cac  at some sites, small palms, etc.) are also recorded in vst_mappingandtagging.

**vst_apparent individual** - For each unique woody individualID recorded in vst_mappingandtagging, technicians create at least one record in vst_apparen ndividual for each eventID. Multiple records per individualID are created for multibole individuals of the following growthForms: small tree, sapling, single shrub, small shrub. To scale up measurements from individuals to the plot-scale, per growthForm measurement areas are linked by plotID and eventID in vst_perplotperyear.

*Shape* appears only to be recorded for shrubs. 

According to the NEON documentation (NEON.DOC.000987), "Species information may not be informative with respect to assigning growth form. For example, Toxicodendron spp. commonly exist as lianas, small shrubs, and shrub groups." For this reason, growth form was not included as a predictor variable. 

```{r merge_veg_structure}
# create output directory if it does not exist 
check_create_dir(out_dir)

# loop through folders of field data with different dates
dirs <- list.dirs(path = main_path )
dirs <-dirs[ grepl("NEON.D17.SJER", dirs) ]


first_loop <- 1 # loop counter
for (woody_path in dirs){
  
  # mapping and tagging table (contains stem locations)
  woody_mapping_path <- paste(woody_path, 
                              list.files(path = woody_path, 
                                         pattern = "mappingandtagging"), 
                              sep="/")
  
  # apparent individual table (contains height and crown diameter)
  woody_individual_path <- paste(woody_path, 
                                 list.files(path = woody_path, 
                                            pattern = "apparentindividual"), 
                                 sep="/")
  
  # load both data tables 
  # calculate mapped UTM locations of plants from distance/azimuth
  woody_mapping <- read.csv(woody_mapping_path)
  woody_individual <- read.csv(woody_individual_path)
  
  # rename the date column in vst_mapping to differentiate it
  # from the "date" column in vst_individual.
  # the latter will be used to assess the most recent entries
  names(woody_mapping)[names(woody_mapping) == 'date'] <- 'date.mapping'
  
  # match mapped stems from "vst_mappingandtagging" with structure data 
  # from "vst_apparentindividual" based on individualID 
  merged <- merge(woody_mapping,
                  woody_individual,
                  by="individualID")
  
  # combine woody veg structure data to a single data frame 
  if (first_loop == 1){
    woody_all <- merged
    woody_mapping_all <- woody_mapping
    woody_individual_all <- woody_individual
    first_loop <- 0
    
  } else {
    woody_all <- rbind(woody_all, merged)
    woody_mapping_all <- rbind(woody_mapping_all, woody_mapping)
    woody_individual_all <- rbind(woody_individual_all, woody_individual)
  }
}

# remove duplicate entries; keep most recent
merged <- woody_all %>% 
  group_by(individualID) %>%
  slice(which.max(as.Date(date)))

# write merged vegetation structure tables to file 
write.csv(merged, file = paste(out_dir,"vst_merged.csv"))

# only keep complete entries with the metrics of interest 
woody <- merged[complete.cases(merged$taxonID) &
                  complete.cases(merged$scientificName) & 
                  complete.cases(merged$height) &  
                  complete.cases(merged$maxCrownDiameter) & 
                  #complete.cases(merged$stemDiameter) & 
                  complete.cases(merged$growthForm),]

# keep only the relevant columns 

```


```{r species_table, eval=FALSE}

# Manual lookup of common names 
manu_df <- data.frame(matrix("", ncol = 0, nrow = 16))

manu_df <- manu_df %>%
  mutate(commonNames = c("Unknown",
                 "California buckeye", #AECA
                 "Mariposa manzanita", #ARVIM
                 "Buckbrush", #CECU
                 "Chaparral whitethorn", #CELE2
                 "Jimsonweed", #DAWR2
                 "California buckthorn", #FRCAC7
                 "Silver lupine", #LUAL4
                 "California foothill pine", #PISA2
                 "Blue oak", #QUDO
                 "Interior live oak", #QUWI2
                 "Buckthorn", #RHAMNA
                 "Hollyleaf redberry", #RHIL
                 "Blue elderberry", #SANIC5
                 "Threadleaf ragwort", #SEFL3
                 "Poison Oak" #TODI
                  ),
         taxonID = c("2PLANT",
                     "AECA",
                     "ARVIM",
                     "CECU",
                     "CELE2",
                     "DAWR2",
                     "FRCAC7",
                     "LUAL4",
                     "PISA2",
                     "QUDO",
                     "QUWI2",
                     "RHAMNA",
                     "RHIL",
                     "SANIC5",
                     "SEFL3",
                     "TODI"))

woody_test <- woody %>%
  mutate(treeORshrub = ifelse(grepl("tree", growthForm), 
                      "tree", "shrub"),
         sciNameShort = word(scientificName, 1, 2)) %>%
  group_by(taxonID, sciNameShort, treeORshrub) %>%
  summarise(total = n()) %>% 
  ungroup() %>%
  inner_join(manu_df, by = "taxonID") %>%
  select(total, taxonID, sciNameShort, commonNames, treeORshrub)

# growthForm %in% c("multi-bole tree", "single bole tree","small tree")

#  mutate(sciNameShort = word(scientificName, 1, 2),
#         commonName = sci2comm(scinames = sciNameShort, db = "itis"))

kable(woody_test)
```


```{r explore_data, eval=FALSE}
# TESTING using table with common names 
ggplot(aes(x = commonNames, y = total, fill = treeORshrub), data = woody_test) + 
  geom_bar(stat="identity") + 
  labs(y = "Total # of plants", x = "Common name") + 
  ggtitle("Number of plants per species") +
  scale_fill_manual(name = "Growth Type",
                      labels = c("shrub", "tree"),
                      values = c("grey", "darkgreen")) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        panel.background = element_blank(),
        panel.grid.major = element_line(color = "grey10",
                                        linetype = 3,
                                        size = .1))

ggplot(data = woody_test, aes(x = taxonID)) +
  geom_boxplot(data = woody, aes(y = height))
```

## Random Forest 

Useful tutorials for the randomForest function in R: 

 - [trevorstephens](http://trevorstephens.com/kaggle-titanic-tutorial/r-part-5-random-forests/) 
 - [kaggle](https://www.kaggle.com/mrisdal/exploring-survival-on-the-titanic)
 - [keyrus blog](http://blog.keyrus.co.uk/alteryxs_r_random_forest_output_explained.html)
 - [Tuning parameters](https://machinelearningmastery.com/tune-machine-learning-algorithms-in-r/)
 - [Variable importance](https://medium.com/the-artificial-impostor/feature-importance-measures-for-tree-models-part-i-47f187c1a2c3)
 - [Detailed description of RF](https://www.stat.berkeley.edu/~breiman/RandomForests/cc_home.htm)

Random Forests grows many classification trees. To classify a new object from an input vector, put the input vector down each of the trees in the forest. Each tree gives a classification, and we say the tree "votes" for that class. The forest chooses the classification having the most votes (over all the trees in the forest).

Using random forest algorithm to classify species of shrubs at the SJER site in central california. How accurately can species be predicted from structural measurements? 

According to the species histogram, some species have very few samples in the data set.The data set will be filtered to maintain only species that have >20 samples. 303 samples across 5 species were used with 3 descriptor variables (shape, height, crown diameter) were used. 

To avoid classifier bias, the number of samples was reduced to 46, the minimum number of samples across the species. in comparison, CECU had 121 samples and if the number of samples per class was not adjusted, this resulted in a great overestimation of CECU by the classifier. 

Tuning parameters for RF: 
- mtry: Number of variables randomly sampled as candidates at each node/split; finds a variable (and value of that variable) which optimizes the split
- ntree: Number of trees to grow. (used the default, 500)

To split the available data into training and test sets, Leave One Out Cross Validation (LOOCV) was used to avoid overfitting in the final model and to ensure that all species were represented in model training. 

Random seed was set for reproducibiliy. 
Using the caret package in R, LOOCV was performed and yielded the following performance metrics: 
  mtry=2 yielded the highest accuracy, 0.39
  
The actual and predicted species for every iteration of the LOOCV were plotted in the confusionMatrix. 
  Overall predictive accuracy: 39%
  Kappa (the amount of agreement correct by the agreement expected by chance. The values range from  1, which indicates complete disagreement between classes, to +1, which indicates a perfect agreement): 0.24
  

It gives estimates of what variables are important in the classification; in other words, which features add the most predictive power to the model. (Sort of like coefficients?) The scales are usually arbitrary, and there are two different criteria used that can generate different importance rankings. 
  
  (1) Gini of Mean Decrease in Impurity (MDI) aka Node Impurity / Mean Decrease Gini
    - difference between RSS before and after the split on that variable
    - Look at each node, which variable was used to split. 
    - GINI importance measures the average gain of purity by splits of a given variable. If the variable is useful, it tends to split mixed labeled nodes into pure single class nodes.
    
  (2) Permutation Importance or Mean Decrease in Accuracy (MDA)
    - MDA is calculated by excluding each variable in the model. A high MDA means that the variable has an important impact on the model’s classification performance.
    - If a variable is assigned values by random permutation (that are present in the data), how much will the MSE increase. Higher value, higher variable importance.

As shown in the variable importance plot, there is a different variable importance ranking depending on the metric. 
  - MaxCrownDiameter is the most important
  - MDA shows that shape and height are the next important
  - MDG shows that height and then shape are the next most important variables

RF uses an Out-of-Bag (OOB) sampling procedure to provide a reliable error rate: Two thirds of the samples were randomly selected and used for training while the remaining third is used for validation. OOB is the mean prediction error on each training sample. 


```{r}
# set the random seed in R for reproducibility
set.seed(14)

# keep species with at least 20 samples 
woody_gt <- woody %>%
                group_by(taxonID) %>% 
                  filter(n()>20)

# further filter data drame for only columns of interest
woody_input2 <- woody_gt %>% 
          select(shape, 
                 height, 
                 maxCrownDiameter, 
                 taxonID) %>% 
          na.omit()

# reduce number of samples per species to avoid classifier bias 
sampleSummary <- woody_input2 %>%
                group_by(taxonID) %>%
                  summarize(total = n())
minSamples <- min(sampleSummary$total)

# isolate the samples per species
taxon1 <- woody_input2[woody_input2$taxonID=="CECU",]
taxon2 <- woody_input2[woody_input2$taxonID=="CELE2",]
taxon3 <- woody_input2[woody_input2$taxonID=="LUAL4",]
taxon4 <- woody_input2[woody_input2$taxonID=="RHIL",]
taxon5 <- woody_input2[woody_input2$taxonID=="TODI",]
# keep random minSamples of each species; merge
species1 <- taxon1[sample(nrow(taxon1), minSamples), ]
species2 <- taxon2[sample(nrow(taxon2), minSamples), ]
species3 <- taxon3[sample(nrow(taxon3), minSamples), ]
species4 <- taxon4[sample(nrow(taxon4), minSamples), ]
species5 <- taxon5[sample(nrow(taxon5), minSamples), ]
input <- rbind(species1, species2, species3, species4, species5)

# convert to categorical variables for classifier 
input$shape <- factor(input$shape)

# LOOCV using caret package. define training control
train_control <- trainControl(method="LOOCV")

# train the model
rf <- train(taxonID~., 
            data=input, 
            trControl=train_control, 
            importance = TRUE,
            method="rf")

# keep mtry=2, since it yields the highest accuracy 0.3882353
# with Kappa value  0.2352941 
rf.2 <- rf$pred[seq(1, nrow(rf$pred), 3), ]

# plot confusion matrix 
cm.2 <- data.frame(melt(xtabs( ~ rf.2$pred + rf.2$obs)))
ggplot(data = cm.2,
       mapping = aes(x = rf.2.pred,
                     y = rf.2.obs)) +
  geom_tile(aes(fill = value)) +
  geom_text(aes(label = sprintf("%1.0f", value)), vjust = 1) +
  scale_fill_gradient(low = "gray",
                      high = "coral1",
                      trans = "log") +
  labs(x = "Predicted", y = "Actual") + 
  theme(legend.position="none")

confusionMatrix(rf.2$pred, rf.2$obs)

# calculate total accuracy, user's, and producers 
o.accuracy <- rf$results$Accuracy[1]

cm.tab <- xtabs(~ rf.2$pred + rf.2$obs)
# overall accuracy
sums <- vector()
# for(i in 1:dim(cm.tab)[1]){
#   sums[i] <- cm.tab[i,i]
# }
# o.accuracy <- sum(sums)/sum(cm.tab)
# User accuracy
u.accuracy <- diag(cm.tab) /rowSums(cm.tab)
# Producer accuracy
p.accuracy <- diag(cm.tab)/colSums(cm.tab)
userProducerAcc <- rbind(u.accuracy, p.accuracy)

# plot variable importance 
varImpPlot(rf$finalModel)

# boxplot of predictor variable ranges per species 
# HEIGHT
b1 <- ggplot(aes(taxonID, height, fill = taxonID), data = input) +
  geom_boxplot() + 
  ggtitle("Height") + 
  labs(x = "Species ID", y = "Value") + 
  scale_fill_discrete(name = "Species ID") + 
  theme(legend.position="none",
        axis.text.x=element_text(angle=45, hjust=1))
# CROWN DIAMETER
b2 <- ggplot(aes(taxonID, maxCrownDiameter, fill = taxonID), data = input) +
  geom_boxplot() + 
  ggtitle("Crown Diameter") + 
  labs(x = "Species ID", y = " ") + 
  scale_fill_discrete(name = "Species ID") + 
  theme(legend.position="none",
        axis.text.x=element_text(angle=45, hjust=1))
grid.arrange(b1, b2, nrow = 1)

```

